---
title: "Simulation 1"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(broom)
library(rsample)
library(ggdag)
library(survey)
library(tableone)
library(tinytex)
library(mvtnorm)
#devtools::install_github('osofr/simcausal', build_vignettes = FALSE)
library(simcausal)
library(rpart)
library(randomForest)
library(gbm)
library(caret)
```

I begin by simulating a fairly simple causal effect model. Let Y denote the response variable, $Z_1, Z_2$ denote covariate predictor variables, and $X$ denote a treatment indicator variable. Let $(Z_1, Z_2), U, V$ be independent, 

Response/predictor relationship: 

$Y = a + bX + c_1 Z_1 + c_2 Z_2 + dU$

Treatment assignment model: 

\begin{equation}
X = 
  \begin{cases}
  1, & \text{if} \ e + f_1Z_1 + f_2Z_2 + V > 0 \\
  0, & \text{else}
  \end{cases}
\end{equation}

```{r}
dag <- DAG.empty() + 
  node(c("Z1", "Z2"), distr = "mvtnorm::rmvnorm", asis.params = list(mean="c(0.5,1)", sigma = "matrix(c(2,1,1,1), ncol=2)")) + 
  node("U", distr = "rnorm", mean = 0, sd = 1) + 
  node("V", distr = "rnorm", mean = 0, sd = 1) + 
  node("X", distr = "rbern", prob = plogis(0.5 + 0.25*Z1 + 0.75*Z2 + V)) + 
  node("Y", distr = "rconst", const = 1 + X + 0.1*Z1 + 2*Z2 + U)

dag <- set.DAG(dag)
sample.size <- 1000
sim1.data <- sim(dag, n=sample.size)
```
In the case of simulations, we know the relationships between variables prior to creating a model. As such, we can visualize the underlying variable relationships through a DAG. 

```{r}
sim1.dag <- dagify(X ~ Z1 + Z2 + V, Y ~ X + Z1 + Z2 + U, exposure = "X", outcome = "Y")

sim1.dag.viz <- sim1.dag %>% 
  ggdag() + 
  labs(title = "Simulation 1 DAG")

```

We begin by visualizing the data. 

```{r}
sim1.data %>% 
  ggplot(aes(Y, fill = factor(X))) + 
  geom_vline(xintercept = 0) + 
  geom_density(color = "white", alpha = 0.75, size = 0.5) + 
  scale_fill_manual(values = c("red", "blue")) + 
  theme_minimal() + 
  labs(x = "Y value", fill = "X value (0 or 1)", title = "Initial Distribution of Treatment Assignment")
```
Now, I will pretend as though I do not know the mechanism behind treatment assigment. I will first use logistic regression to estimate the probability of treatment assignment. 

```{r}
# logistic regression
logistic.regression <- glm(X ~ Z1 + I(Z1^2) + Z2 + I(Z2^2) + V + I(V^2), 
                           family = "binomial", 
                           data = sim1.data)
sim1.data <- sim1.data %>% 
  mutate(logreg.fitted = logistic.regression$fitted.values)
```

Now, I will use a single decision tree to model the probability of treatment. 
```{r CART}
regression.tree <- rpart(X ~ Z1 + I(Z1^2) + Z2 + I(Z2^2) + V + I(V^2), data = sim1.data, method = "class")
predictions <- predict(regression.tree, type = "prob")[,2]

sim1.data <- sim1.data %>% 
  mutate(regtree.fitted = predictions)
```

Now, I model the probability of treatment using a random forest. 
```{r random forest}
preds <- sim1.data[,c(2,3,5)]
resp <- as.factor(sim1.data[,6])

tune.rf <- tuneRF(preds, resp, stepFactor = 2, ntreeTry = 500)
# optimal: mtry = 1

random.forest <- randomForest(as.factor(X) ~ Z1 + I(Z1^2) + Z2 + I(Z2^2) + V + I(V^2), data = sim1.data, ntrees = 500, mtry = 1)

rf.predictions <- predict(random.forest, type = "prob")[,2]

sim1.data <- sim1.data %>% 
  mutate(rf.fitted = rf.predictions)
```


```{r gbm}
fitcontrol <- trainControl(method = "repeatedcv", 
                           number = 10, 
                           repeats = 10)
gbm <- train(as.factor(X) ~ Z1 + Z2 + V, 
             data = sim1.data, 
             method = "gbm", 
             trControl = fitcontrol, 
             verbose =FALSE)

sim1.data <- sim1.data %>% 
  mutate(gbm.fitted = predict(gbm, type = "prob")[,2])

```
Now, I will calculate the initial bias, post-stratification bias, and resulting bias reduction. 
```{r}
# initial bias, prior to subclassification
b1.regtree <- mean(sim1.data$logreg.fitted[sim1.data$X == 1]) - mean(sim1.data$logreg.fitted[sim1.data$X == 0])
b1.regtree <- mean(sim1.data$regtree.fitted[sim1.data$X == 1]) - mean(sim1.data$regtree.fitted[sim1.data$X == 0])
b1.rf <- mean(sim1.data$rf.fitted[sim1.data$X == 1]) - mean(sim1.data$rf.fitted[sim1.data$X == 0])
b1.gbm <- mean(sim1.data$gbm.fitted[sim1.data$X == 1]) - mean(sim1.data$gbm.fitted[sim1.data$X == 0])

# create strata and assign each observation 
sim1.data <- sim1.data %>% 
  mutate(decile.logreg = cut(logreg.fitted, breaks = 10, labels = c("1", "2", "3", "4", "5", "6", "7", "8", "9", "10")), 
         decile.regtree = cut(regtree.fitted, breaks = 10, labels = c("1", "2", "3", "4", "5", "6", "7", "8", "9", "10")), 
         decile.rf = cut(rf.fitted, breaks = 10, labels = c("1", "2", "3", "4", "5", "6", "7", "8", "9", "10")),
         decile.gbm = cut(gbm.fitted, breaks = 10, labels = c("1", "2", "3", "4", "5", "6", "7", "8", "9", "10")))

# calculate post-stratification bias 
expect.decile.logreg <- seq(1,10, by = 1)
expect.decile.regtree <- seq(1,10, by = 1)
expect.decile.rf <- seq(1,10, by = 1)
expect.decile.gbm <- seq(1,10, by = 1)

TE.logreg <- rep(0, 10)
CE.logreg <- rep(0,10)
TE.regtree <- rep(0,10)
CE.regtree <- rep(0,10)
TE.rf <- rep(0,10)
CE.rf <- rep(0,10)
TE.gbm <- rep(0,10)
CE.gbm <- rep(0,10)


for(i in 1:length(expect.decile.logreg)){
  
  expect.decile.logreg[i] <- ifelse(length(sim1.data$logreg.fitted[sim1.data$X == 1 & sim1.data$decile.logreg == i]) == 0 || length(sim1.data$logreg.fitted[sim1.data$X == 0 & sim1.data$decile.logreg == i]) == 0, 0, mean(sim1.data$logreg.fitted[sim1.data$X == 1 & sim1.data$decile.logreg == i]) - mean(sim1.data$logreg.fitted[sim1.data$X == 0 & sim1.data$decile.logreg == i]))
  
  TE.logreg[i] <- ifelse(length(sim1.data$logreg.fitted[sim1.data$X == 1 & sim1.data$decile.logreg == i]) == 0, 0, mean(sim1.data$Y[sim1.data$X == 1 & sim1.data$decile.logreg == i]))
  
  CE.logreg[i] <- ifelse(length(sim1.data$logreg.fitted[sim1.data$X == 0 & sim1.data$decile.logreg == i]) == 0, 0, mean(sim1.data$Y[sim1.data$X == 0 & sim1.data$decile.logreg == i]))
  
  expect.decile.regtree[i] <- ifelse(length(sim1.data$regtree.fitted[sim1.data$X == 1 & sim1.data$decile.regtree == i]) == 0 || length(sim1.data$regtree.fitted[sim1.data$X == 0 & sim1.data$decile.regtree == i]) == 0, 0, mean(sim1.data$regtree.fitted[sim1.data$X == 1 & sim1.data$decile.regtree == i]) - mean(sim1.data$regtree.fitted[sim1.data$X == 0 & sim1.data$decile.regtree == i]))
  
  TE.regtree[i] <- ifelse(length(sim1.data$regtree.fitted[sim1.data$X == 1 & sim1.data$decile.regtree == i]) == 0, 0, mean(sim1.data$Y[sim1.data$X == 1 & sim1.data$decile.regtree == i]))
  
  CE.regtree[i] <- ifelse(length(sim1.data$regtree.fitted[sim1.data$X == 0 & sim1.data$decile.regtree == i]) == 0, 0, mean(sim1.data$Y[sim1.data$X == 0 & sim1.data$decile.regtree == i]))
  
  expect.decile.rf[i] <- ifelse(length(sim1.data$rf.fitted[sim1.data$X == 1 & sim1.data$decile.rf == i]) == 0 || length(sim1.data$rf.fitted[sim1.data$X == 0 & sim1.data$decile.rf == i]) == 0, 0, mean(sim1.data$rf.fitted[sim1.data$X == 1 & sim1.data$decile.rf == i]) - mean(sim1.data$rf.fitted[sim1.data$X == 0 & sim1.data$decile.rf == i]))
  
  TE.rf[i] <- ifelse(length(sim1.data$rf.fitted[sim1.data$X == 1 & sim1.data$decile.rf == i]) == 0, 0, mean(sim1.data$Y[sim1.data$X == 1 & sim1.data$decile.rf == i]))
  
  CE.rf[i] <- ifelse(length(sim1.data$rf.fitted[sim1.data$X == 0 & sim1.data$decile.rf == i]) == 0, 0, mean(sim1.data$Y[sim1.data$X == 0 & sim1.data$decile.rf == i]))
  
  expect.decile.gbm[i] <- ifelse(length(sim1.data$gbm.fitted[sim1.data$X == 1 & sim1.data$decile.gbm == i]) == 0 || length(sim1.data$gbm.fitted[sim1.data$X == 0 & sim1.data$decile.gbm == i]) == 0, 0, mean(sim1.data$gbm.fitted[sim1.data$X == 1 & sim1.data$decile.gbm == i]) - mean(sim1.data$gbm.fitted[sim1.data$X == 0 & sim1.data$decile.gbm == i]))
  
  TE.gbm[i] <- ifelse(length(sim1.data$gbm.fitted[sim1.data$X == 1 & sim1.data$decile.gbm == i]) == 0, 0, mean(sim1.data$Y[sim1.data$X == 1 & sim1.data$decile.gbm == i]))
  
  CE.gbm[i] <- ifelse(length(sim1.data$gbm.fitted[sim1.data$X == 0 & sim1.data$decile.gbm == i]) == 0, 0, mean(sim1.data$Y[sim1.data$X == 0 & sim1.data$decile.gbm == i]))
  
}


b2.logreg <- sum(expect.decile.logreg)
b2.regtree <- sum(expect.decile.regtree)
b2.rf <- sum(expect.decile.rf)
b2.gbm <- sum(expect.decile.gbm)

reduction.logreg <- (1 - b2.logreg/b1.logreg)*100
reduction.regtree <- (1 - b2.regtree/b1.regtree)*100 
reduction.rf <- (1 - b2.rf/b1.rf)*100
reduction.gbm <- (1 - b2.gbm/b1.gbm)*100

# count obs by decile in treatment vs. control group 
treatment.nobs.logreg <- sim1.data %>% 
  filter(X==1) %>% 
  group_by(decile.logreg) %>% 
  count()

control.nobs.logreg <- sim1.data %>% 
  filter(X == 0) %>% 
  group_by(decile.logreg) %>% 
  count()

treatment.nobs.regtree <- sim1.data %>% 
  filter(X==1) %>% 
  group_by(decile.regtree) %>% 
  count()

control.nobs.regtree <- sim1.data %>% 
  filter(X==0) %>% 
  group_by(decile.regtree) %>% 
  count()

treatment.nobs.rf <- sim1.data %>% 
  filter(X==1) %>% 
  group_by(decile.rf) %>% 
  count()

control.nobs.rf <- sim1.data %>% 
  filter(X==0) %>% 
  group_by(decile.rf) %>% 
  count()

treatment.nobs.gbm <- sim1.data %>% 
  filter(X==1) %>% 
  group_by(decile.gbm) %>% 
  count()

control.nobs.gbm <- sim1.data %>% 
  filter(X==0) %>% 
  group_by(decile.gbm) %>% 
  count()
```

Now, I will focus on calculating the Average Treatment Effect. 

```{r}
# add the first decile to df
row <- data.frame(1,0)
names(row) <- c("decile.logreg", "n")
treatment.nobs.logreg <- treatment.nobs.logreg[-2,]
treatment.nobs.logreg <- rbind(row, treatment.nobs.logreg)

row2 <- data.frame(1,0)
names(row2) <- c("decile.gbm", "n")
treatment.nobs.gbm <- rbind(row2, treatment.nobs.gbm)

ATE.logreg <- sum((treatment.nobs.logreg[,2] + control.nobs.logreg[,2])*(TE.logreg - CE.logreg))/sample.size

ATE.regtree <- sum((treatment.nobs.regtree[,2] + control.nobs.regtree[,2])*(TE.regtree - CE.regtree))/sample.size

ATE.rf <- sum((treatment.nobs.rf[,2] + control.nobs.rf[,2])*(TE.rf - CE.rf))/sample.size

ATE.gbm <- sum((treatment.nobs.gbm[,2] + control.nobs.gbm[,2])*(TE.gbm - CE.gbm))/sample.size
```
It seems as though a single regression tree is able to extract the causal relationship most effectively in terms of both bias reduction and accuractly of treatment effect. 


